Understanding Implications of Dataset Choice for Feature Effect Estimation: A Simulation-Based Investigation through Error Decomposition

### 2025-02-03 paper/paper.tex

\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. We propose the following estimators:
\begin{equation}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
\end{equation}
\begin{equation}
    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
\end{equation}
\begin{equation}
    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
    \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
\end{equation}

@@

@@

\paragraph{Model Inducers.}
We consider the following model inducers for our simulation study:
\begin{itemize}[label=--]
    \item \textbf{LinReg}: a simple linear regression model as baseline.
    \item \textbf{GAM\_OT}: a Generalized Additive Model (GAM) with spline terms
          and tensor splines for first-order interactions, number of splines
          and penalization optimally tuned.
    \item \textbf{GAM\_OT}: as above, but with hyperparameters chosen to overfit.
    \item \textbf{SVM\_OT}: a Support Vector Machine (SVM) with optimally tuned hyperparameters.
    \item \textbf{SVM\_OF}: a Support Vector Machine (SVM) with hyperparameters chosen to overfit.
    \item \textbf{XGBoost\_OT}: an XGBoost model with optimally tuned hyperparameters.
    \item \textbf{XGBoost\_OF}: an XGBoost model with hyperparameters chosen to overfit.
\end{itemize}

\noindent Hyperparameters are pre-selected, i.e.\ carefully hand-picked for overfitting scenarios
and tuned on a different sample for optimal tuning. Details on the hyperparameters and the
performance of the models can be found in \textsc{Appendix~\ref{app:model-hps-perf}}.

@@

\begin{figure}
    \includegraphics[width=\textwidth]{img/model_performances.png}
    \caption{Performance of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
    \label{fig:model-performance}
\end{figure}

>>>>

\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. We propose the following estimators:
\begin{equation}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
    \label{eq:mse-estimator}
\end{equation}
\begin{equation}
    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
    \label{eq:bias-estimator}
\end{equation}
\begin{equation}
    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
    \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
    \label{eq:variance-estimator}
\end{equation}

@@

To address the research question of which data to choose to estimate feature
effects, we conduct a comprehensive simulation study. For different models,
datasets, and dataset sizes, we estimate the feature effects PDP and ALE on
training data, validation data, and in a cross-validated manner. We then
compute the feature effect error as MSE with respect to the ground truth
feature effects and decompose it into bias and variance components. By
comparing these error terms across the different estimation strategies
(training, validation, cross-validation), we aim to provide nuanced empirical
insights into the implications of the data choice on feature effect estimation.
For deeper analysis, we conduct two ablation studies: one decomposing the
feature effect variance into model variance and Monte Carlo variance, and
another examining the impact of the dataset size on the Monte Carlo variance.
In the following, we describe our experimental set-up for the main simulation
study (\textsc{Section~\ref{sec:experiment-main}}) and the ablation studies
(\textsc{Section~\ref{sec:experiment-ablation}}).

\subsection{Main Experiment}\label{sec:experiment-main}

@@

\paragraph{Models.}
We consider a set of seven complementary learners, spanning different modeling
paradigms:

\begin{itemize}[label=--]
    \item \textbf{LinReg}: a simple linear regression model as a baseline.
    \item \textbf{GAM\_OT}: a Generalized Additive Model (GAM) with spline terms
          and tensor splines for first-order interactions, number of splines
          and penalization optimally tuned.
    \item \textbf{GAM\_OF}: as above, but with hyperparameters chosen to overfit.
    \item \textbf{SVM\_OT}: a Support Vector Machine (SVM) with optimally tuned hyperparameters.
    \item \textbf{SVM\_OF}: a Support Vector Machine (SVM) with hyperparameters chosen to overfit.
    \item \textbf{XGBoost\_OT}: an XGBoost model with optimally tuned hyperparameters.
    \item \textbf{XGBoost\_OF}: an XGBoost model with hyperparameters chosen to overfit.
\end{itemize}

\noindent Hyperparameters are pre-selected, i.e.\ carefully hand-picked for overfitting scenarios
and tuned on a different dataset for optimal tuning. Details on the hyperparameters and performances
of the models can be found in \textsc{Appendix~\ref{app:model-hps-perf}}.

\paragraph{Feature Effect Estimation.}
For the trained models, we estimate the feature effects 
$\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
Monte Carlo integration, we use for each the full training and validation set
as well as a cross-validation strategy. In the latter case, we use a 5-fold
cross-validation on the combined samples of training and validation set, in
each fold fit the model on four folds, and estimate the feature effects on the
remaining fold. We then average the feature effects over all folds. We also
estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
analytical computation infeasible. For the Monte Carlo integration, we use
10000 samples additionally sampled from the DGP, which introduces only a
negligible additional variance term as we show in (\dots) We use 100 grid
points, defined by the quantiles of the theoretical distribution of $X_S$ to
enable a comparison across feature effects. Additionally, we center the curves
after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
omit the first and the last grid point after estimation to avoid boundary
effects especially occuring in ALE plots.

\paragraph{Feature Effect Errors.} To quantify the error of the estimated feature effects, we compute the MSE of
the estimated model feature effects with respect to the estimated ground truth
features effects, as well as their bias and variance. For that, we use the
estimators defined in equations~(\ref{eq:mse-estimator}) to
(\ref{eq:variance-estimator}), using $\widehat{PDP}_{f,S}$ instead of
$PDP_{f,S}$ as ground truth (analogously for ALE). We repeat each
dataset-size-model combination $M=30$ times to estimate the error terms, where
each repetition involves drawing a new training and validation set from the
DGP, refitting the models, and estimating the feature effects. To aggregate the
errors into a single measure per feature effect, we average the errors over the
marginal distribution of $X_S$ by averaging over the grid points.

\subsection{Ablation Experiments}\label{sec:experiment-ablation}
\dots

@@

\begin{figure}
    \includegraphics[width=\textwidth]{img/model_performances.png}
    \caption{R2-score of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
    \label{fig:model-performance}
\end{figure}
